When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-12 doc/paper/vpd_et_paper.tex
back tow ork, jsut schanged a couple citations

This conflicting response between increased demand and higher stomatal stress leads to two potential but opposing perspectives on how ET adjusts to changes in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies (CITATION NEEDED), typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion (Swann et al. 2016 FOR EXAMPLE). On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness I order to save \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. Therefore, an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constants, the slope parameter ($g_1$), and $uWUE$, both reflecting water conservation strategy. The slope parameter is a leaf-scale term related to the willingness of stomata to trade water for CO_2 and to keep stomata open. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). In the eddy-covariance FLUXNET2015 data described in Section \ref{data}, every term in Equation \ref{et} is observed, except for the two parameters $g_1$, and $uWUE$, which have however been measured for the PFTs considered here (REF NEEDED HERE). However, eddy-covariance data are inherently noisy so that we include a measure of uncertainty in our analysis. To account for both mean deviations of $uWUE$ and uncertainty, we introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only very weakly on climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation (De Schepper, V., T. De Swaef, I. Bauweraerts, and K. Steppe (2013), Phloem transport: a review of mechanisms and controls, J Exp Bot, 64(16), 4839–4850, doi:10.1093/jxb/ert302.
Nikinmaa, E., T. Hölttä, P. Hari, P. Kolari, A. Makela, S. Sevanto, and T. Vesala (2012), Assimilate transport in phloem sets conditions for leaf gas exchange, Plant Cell Environ, 36(3), 655–669, doi:10.1111/pce.12004.
Ryan, M. G., and S. Asao (2014), Phloem transport in trees, Tree Physiol, 34(1), 1–4, doi:10.1093/treephys/tpt123.
).

@@

% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.
%\bibliography{references.bib}
\input{vpd_et_paper_PG_v1.bbl}

>>>>

This conflicting response between increased demand and higher stomatal stress leads to two potential but opposing perspectives on how ET adjusts to changes in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness I order to save \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. Therefore, an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constants, the slope parameter ($g_1$), and $uWUE$, both reflecting water conservation strategy. The slope parameter is a leaf-scale term related to the willingness of stomata to trade water for CO$_2$ and to keep stomata open. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). In the eddy-covariance FLUXNET2015 data described in Section \ref{data}, every term in Equation \ref{et} is observed, except for the two parameters $g_1$, and $uWUE$, which have however been measured for the PFTs considered here (REF NEEDED HERE). However, eddy-covariance data are inherently noisy so that we include a measure of uncertainty in our analysis. To account for both mean deviations of $uWUE$ and uncertainty, we introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation (De Schepper, V., T. De Swaef, I. Bauweraerts, and K. Steppe (2013), Phloem transport: a review of mechanisms and controls, J Exp Bot, 64(16), 4839–4850, doi:10.1093/jxb/ert302.
Nikinmaa, E., T. Hölttä, P. Hari, P. Kolari, A. Makela, S. Sevanto, and T. Vesala (2012), Assimilate transport in phloem sets conditions for leaf gas exchange, Plant Cell Environ, 36(3), 655–669, doi:10.1111/pce.12004.
Ryan, M. G., and S. Asao (2014), Phloem transport in trees, Tree Physiol, 34(1), 1–4, doi:10.1093/treephys/tpt123.
).

@@

% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.
\bibliography{references.bib}
% \input{vpd_et_paper.bbl}
